---
title: "Michigan Ecosystem Explorer"
output: 
  flexdashboard::flex_dashboard:
    storyboard: true
---

```{r setup, include=FALSE}
library(flexdashboard)
```


### Welcome! {data-commentary-width=500}

```{r}
library(tmap)
library(sf)
library(tidyverse)
library(raster)

shp <- st_read("./us_state_bounds/tl_2022_us_state.shp", quiet = T) %>%
  filter(NAME == "Michigan")
tmap_mode("view")
tm_shape(shp) +
  tm_borders(col = "gray", lwd = 4) +
  tm_basemap("Esri.WorldImagery")

```

***
<font size = "4"> Dashboard by Myles Walimaa and Randy Swaty </font>  

This idea was initially designed by [Randy Swaty](mailto:rswaty@tnc.org) as an alternative to a PowerPoint presentation. There was interest in building on this idea in the conservation community and I, [Myles Walimaa](mailto:mwalimaa@gmail.com), made some updates with newer data and new charts. This can certainly be built upon and these workflows may be used for any area of interest. The State of Michigan is the target area for this dashboard, since it is where both Randy and I have lived for years.  

Randy is an ecologist for The Nature Conservancy and I am a former National Park Service employee. We both work with the [Conservation Data Lab](https://conservationdatalab.org/) to mentor students in the art of computer coding and data visualization, particularly for conservation purposes.  

The purpose of this dashboard is to explore the major ecosystems of Michigan's past and how they have changed.  
As you explore, ask yourself some of these questions:  

* What is an "ecosystem" to you? How coarse or fine should we define these?  
* What percentage of Michigan was/is covered by wetlands? How important are wetlands to you? What is their ecological value?  
* How often do you think wildfires occurred historically? Can you think of any ecosystems that are fire dependent?  
* Which ecosystems in Michigan have you seen? Are there any you would like to see?  
* Are there any trends that you have noticed, based off the data presented?

<font size = "3"> *PLEASE read the disclaimers on each slide explaining the assumptions with each dataset. This is meant for exploratory purposes only.*</font>  

**Michigan**  

- Michigan is located in America's "midwest" region, one of the more northern states.  

- It shares borders with Ohio, Indiana, Illinois (nautical), and Wisconsin, as well as international borders with Canada.  

- It contains two peninsulas (upper and lower) separated by the Mackinac Bridge. Residents north of the bridge are called "Yoopers", and residents south/below the bridge are called "Trolls".  

- It is also called the "Great Lakes State", as it borders 4/5 of the Laurentian Great Lakes.  

- It is home to Detroit, the Motor City, as well as the birthplace of [Motown](https://www.youtube.com/watch?v=8Ff2Z0Dq2sg).  

- There are 62 million acres of land within its political borders.  

- Isle Royale National Park is Michigan's only National Park (proper) and is the least visited park in the NPS system, as it takes a 5 hour ferry ride to get to the incredibly remote island.  


### Looking Back {data-commentary-width=500}

```{r}


library(tidyverse)
options(scipen = 999)

bps_att3 <- read.csv("./outputs/bps_att_chart.csv") %>%
  top_n(25)

chart2 <- ggplot(data = bps_att3, aes(fill = GROUPVEG, y = acres, x = reorder(BPS_NAME, acres))) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top 25 Historic Ecosystems in Michigan",
       x = "",
       y = "Acres",
       fill = "Ecosystem Type") +
  theme_light() +
  scale_fill_manual(values = c("#117733", "#cc6677", "#119d74", "#999933", "#25bfde")) +
  scale_y_continuous(labels = scales::comma)

ichart2 <- plotly::ggplotly(chart2) %>% plotly::style(hoverinfo = "none")
ichart2


```


***
**DISCLAIMER:**  
This map was created using LANDFIRE's BioPhysical Settings. As there was no satellite imagery 300 years ago, this is a *model* of what the historical ecosystems looked like based soil types, climate, elevation, ecological succession, among many other datasets.  

- This was intentionally non-interactive simply to give you a broad visual of what things probably looked like.  

- This is coarse ecosystem representation since it is difficult to display several dozen ecosystem types on a single map.  

- The original data displayed plots of 30m x 30m squares that we call "pixels", and Michigan has over 500 million of them!  

- This map has much lower pixel resolution than mentioned above, again to simply give you an idea of what things may have looked like without going into great detail.  

This is what we think the distribution of ecosystems looked like prior to European settlement. The caretakers of this land at the time were the indigenous natives who called the Great Lakes their home. They recognized the value of fire for regeneration and resiliency, often performing "prescribed" burns to keep their home strong and healthy.  

- Can you think of non-anthroprogenic disturbances to these ecosystems?  

Explore this map created by [Native Land Digital](https://native-land.ca/) to learn more about which tribes were located where. Botanist Robin Wall Kimmerer's book [Braiding Sweetgrass](https://milkweed.org/book/braiding-sweetgrass) provides valuable insight on the traditional ways the natives connected with the land, largely through combining ancient philosophy and modern science.

**DISCLAIMER:**  
This chart was created using LANDFIRE's BioPhysical Settings. As there was no satellite imagery 300 years ago, this is a *model* of what the historical ecosystems looked like based soil types, climate, elevation, ecological succession, among many other datasets.  

- BioPhysical Settings (BPS) have a "fine" scale classification called "BPS_NAME" and a "coarse" scale classification called "GROUPVEG".

- The acreages were calculated based on the quantity of 900m^2^ pixels (30m x 30m), which was converted to acres.  

- When you hover, you can see how many acres of each ecosystem used to exist (roughly).  

- Use your mouse to zoom in on smaller areas by drawing a box.  

- You can reset the zoom by clicking the "house" icon at the top left of the chart.  

How accurate do you think this chart is?  

Do you think BPS_NAME or GROUPVEG has higher accuracy in describing ecosystems?  

Take "North-Central Interior Beech-Maple Forest" for example:  

- GROUPVEG says this is mostly hardwoods, would you agree?  
- Do you think these pixels contain *only* Beech and Maple trees?  
- What is the ratio of Beech-to-Maple in these forests? Do you think it varies?  


The point is that this is *not* a perfect representation, although it helps us gain a general understanding of what probably was going on several hundred years ago. If you want to explore the BPS dataset further, you can find [information here](https://www.landfire.gov/bps.php) and **Randy knows where there is a .pdf or something that has descriptions of what all the silly BPS_NAMEs mean, with examples of what kind of things will probably be found there... right??? LIKE THE EVT ONE??? IT MUST EXIST SOMEWHERE**



### Visual Exploration of Change {data-commentary-width=500}

```{r}
library(tidyverse)
c2 <- read.csv("./outputs/bps2scl.csv")


c3 <- c2  %>%
  group_by(BPS_NAME, SCL_V) %>%
  summarise(freq = sum(freq)) %>%
  as.data.frame() %>%
  mutate(SCL_V = replace(SCL_V, str_detect(SCL_V, "Ag"), "Agricultural")) %>%
  mutate(acres = as.integer(round(freq * 900 / 4046.85)),
         SCL_V = factor(c(SCL_V), levels= c("Agricultural", "Developed", "Other", "NatVeg")))

c25 <- c3 %>% group_by(BPS_NAME) %>% summarise(total = sum(acres)) %>%
  arrange(desc(total)) %>% top_n(25) %>% pull(BPS_NAME)

c4 <- c3 %>% filter(BPS_NAME %in% c25)


chart <- ggplot(data = c4, aes(fill = SCL_V, y = acres, x = reorder(BPS_NAME, acres))) +
  geom_bar(position = "stack", stat = "identity") +
  coord_flip() +
  scale_fill_manual(values = c("#e2ca1b", "#515151", "#cd7220", "#117733")) +
  labs(title = "Top 25 Historic Ecosystems and How They Changed",
       x = "",
       y = "Acres",
       fill = "Change Type") +
  scale_y_continuous(labels = scales::comma) +
  theme_light()
ichart <- plotly::ggplotly(chart) %>% plotly::style(hoverinfo = "none")
ichart
```

***

**DISCLAIMER:**
The assumptions made in this chart are that *if* the ecosystems are no longer following the trajectory of natural ecological succession, *then* they have been converted into something unnatural. This chart does not consider ecosystems changing from one to another as that may happen naturally, only change into non-native, mineland, agriculture, development, or the like.  

- You can see which of the ecosystems remained relatively untouched, as well as the ones that were heavily affected.  

- The acreages were calculated based on the quantity of 900m^2^ pixels (30m x 30m), which was converted to acres.  

- When you hover, you can see how many acres of each ecosystem changed (roughly). 

- Use your mouse to zoom in on smaller areas by drawing a box.  

- You can reset the zoom by clicking the "house" icon at the top left of the chart.  

Which ecosystems remain mostly the same? Is this surprising?  

Where did most of the agricultural land come from?  Do you notice any trends?  

Are there any ecosystems that might be close to disappearing completely?



**DISCLAIMER:**
The assumptions made in this map are that *if* the ecosystems are no longer following the trajectory of natural ecological succession, *then* they have been converted into something unnatural. This map does not consider ecosystems changing from one to another as that may happen naturally, only change into non-native, mineland, agriculture, development, or the like.  


- 50.1% of ecosystems remain Natural Vegetation  

- 25.9% of ecosystems have been converted to Agriculture  

- 13.0% of ecosystems have been Developed in one way or another  

- 11.0% of ecosystems have changed in Other ways  

Where has most of the ecosystem conversion taken place?  

To what category have most of the ecosystems converted? What used to be there?  


### Coarse Look at Conversion {data-commentary-width=500}

```{r}
library(tidyverse)
c <- read.csv("./outputs/bps2evt.csv")

# finding GROUPVEG categories
group <- c %>% group_by(GROUPVEG) %>% 
  summarise(freq = sum(Freq.x)) %>%
  arrange(desc(freq)) %>%
  pull(GROUPVEG)

# grouping combine and filter
data <- c %>% dplyr::select(GROUPVEG, EVT_PHYS, Freq.x, Freq.y, freq) %>% 
  mutate(EVT_PHYS = replace(EVT_PHYS, str_detect(EVT_PHYS, "Developed"), "Developed")) %>%
  mutate(EVT_PHYS = replace(EVT_PHYS, str_detect(EVT_PHYS, "Exotic"), "Exotic")) %>%
  mutate(EVT_PHYS = replace(EVT_PHYS, str_detect(EVT_PHYS, "Quarries"), "Mineland")) %>%
  group_by(GROUPVEG, EVT_PHYS) %>%
  summarise(freq2 = sum(freq)) %>%
  filter(GROUPVEG %in% group) %>%
  mutate(ACRES = round(freq2 * 900 / 4046.86)) %>%
  rename(source = GROUPVEG,
         target = EVT_PHYS,
         value = ACRES) %>%
  mutate(source = paste0(source, " (past)"),
         target = paste0(target, " (present)"))

##############
##############
# BUILDING SANKEY

nodes <- data.frame(
  name = c(as.character(data$source),
           as.character(data$target)) %>% unique())

data$IDsource <- match(data$source, nodes$name) - 1
data$IDtarget <- match(data$target, nodes$name) - 1

cols <- "d3.scaleOrdinal() .domain(['Conifer (past)', 'Grassland (past)', 'Hardwood (past)', 'Hardwood-Conifer (past)', 'Riparian (past)', 'Agricultural (present)', 'Conifer (present)', 'Conifer-Hardwood (present)', 'Developed (present)', 'Exotic (present)', 'Grassland (present)', 'Mineland (present)', 'Riparian (present)', 'Shrubland (present)']) .range(['#cc6677', '119d74', '#999933', '#25bfde', '#e2ca1b', '#999933', '#515151', 'cd7220', '#cd7220', '#cd7220', '#cd7220', 'red', 'yellow', 'blue', '#117733'])"

snky <- networkD3::sankeyNetwork(Links = data,
                                 Nodes = nodes,
                                 Source = "IDsource",
                                 Target = "IDtarget",
                                 Value = "value",
                                 NodeID = "name",
                                 fontSize = 14,
                                 iterations = 0,
                                 sinksRight = F,
                                 width = 6,
                                 colourScale = cols)
snky
```

***
**DISCLAIMER:**  
Unfortunately, this is *not* an "apples to apples" comparison. The datasets used were BioPhysical Settings (historical model) and Existing Vegetation Type (current satellite imagery, ground assessment, other methods). Additionally, the Existing Vegetation Type is not a perfect representation of the ground because there can be multiple ecosystems found in 900 m^2^ (remember 30m x 30m pixels).  

- This chart provides a general idea of how Michigan land has changed on a coarse scale.  

- BPS and EVT were stacked on top of each other, and we extracted the coarse categories from those datasets.  

- The left side shows coarse landcover from the past (BPS), and the right shows current landcover (EVT).  

- The grey bands represent transitions from one to another.  

- When you hover, the number value is in acres (approx).  

Does anything surprise you about how things have changed?  

Which ecosystems had the least amount of conversion? What about the most?

Why did so little Agricultural land come from Coniferous ecosystems?



### Fine Look at Conversion {data-commentary-width=500}

```{r}
library(tidyverse)
t <- read.csv("./outputs/bps2evt.csv")
bps_att <- read.csv("./outputs/bps_michigan.csv")
evt_att <- read.csv("./outputs/evt_michigan.csv")

t1 <- t %>%
  group_by(BPS_NAME, EVT_NAME, GROUPVEG, EVT_PHYS) %>%
  summarise(freq = sum(freq)) %>%
  mutate(ACRES = round(freq * 900 / 4046.86)) %>%
  dplyr::select(BPS_NAME, GROUPVEG, ACRES, EVT_PHYS, EVT_NAME) %>%
  arrange(desc(ACRES))
            
#t2 <- reactable::reactable(t1, defaultPageSize = 15, filterable = T, searchable = T)
t2 <- DT::datatable(t1, extensions = "Buttons", options = list(paging = T,
                                                               scrollX = T,
                                                               lengthMenu = c(15, 30, 50, 100),
                                                               pageLength = 15,
                                                               autoWidth = T,
                                                               searching = T,
                                                               ordering = T,
                                                               dom = "Bfrtip",
                                                               buttons = c("csv", "excel"))) %>%
  DT::formatCurrency("ACRES", currency = "", interval = 3, mark = ",", digits = 0)

t2
```

***

**DISCLAIMER:**  
Unfortunately, this is *not* an "apples to apples" comparison. The datasets used were BioPhysical Settings (historical model) and Existing Vegetation Type (current satellite imagery, ground assessment, other methods). Additionally, the Existing Vegetation Type is not a perfect representation of the ground because there can be multiple ecosystems found in 900 m^2^ (remember 30m x 30m pixels).  

- For those of you who want to explore this conversion further, search no more!  

- BPS and EVT were overlaid on top of each other the attributes of each pixel were extracted.  

- The left side shows the historical BPS ecosystem, and the right side shows the current EVT ecosystem.  

- The acre value shows how many acres converted from that BPS type to the EVT type.

- There are over 2000 comparisons over 152 pages.  

- The acreages were calculated based on the quantity of 900m^2^ pixels (30m x 30m), which was converted to acres.  

- You can sort columns by ascending or descending order, as well as filter results or search for something specific.  

For example, 429,625 acres of North-Central Oak Barrens are now Eastern Cool Temperate Row Crop. Well, roughly.

Again, this is not a perfect comparison and is only intended on giving you a general idea of what's going on.  


**ENJOY AND HAVE FUN!!**


### Learn More {data-commentary-width=500}

```{r}

```

Sources, LF documentation, ideas, idk some useful stuff to keep pushing these people to research stuff